---
title: "College Football Games and Crimes"
author: "Nabin Oli & Nava Nizard"
date: "`r Sys.Date()`"
output: 
    html_document:
        toc: yes
        toc_depth: 3
    pdf_document:
        toc: yes
        toc_depth: 3
        latex_engine: pdflatex
    header-includes:
        - \usepackage{float}
        - \floatplacement{table}{H}
        - \setlength{\parskip}{6pt}
        - \setlength{\parindent}{0pt}
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  message = FALSE,
  warning = FALSE,
  echo = FALSE
)

library(modelsummary)
library(MASS) # for negative binomial model
library(tidyverse)
library(lubridate)
library(stringr)
library(knitr)

final_table <- read_csv("data/final_table.csv")
```

## Table 2: Distribution of Game Days by Day of the Week
```{r replicate-table-2}
# Calculate number of games per day of the week
num_of_games_per_dotw <- final_table %>% 
    mutate(day_of_the_week = wday(day, label = TRUE)) %>%
    filter(game_status != "no_game") %>%
    group_by(day_of_the_week) %>%
    summarize(num_of_games = n_distinct(game_id))

# Calculate number of {DOTW} observered over the cfb season over 2000-2005 for all ORIs
num_of_observations_per_dotw <- final_table %>% # about 13 (num of {DOTW}) * 6 (years) * 26 (ORIs / num of agencies)
    mutate(day_of_the_week = wday(day, label = TRUE)) %>%
    group_by(day_of_the_week) %>%
    summarize(num_of_observations = n())
    
# Merge results
joined_tbl <- full_join(num_of_games_per_dotw, num_of_observations_per_dotw, by = "day_of_the_week")

# Add totals row
totals_row <- tibble(
  day_of_the_week = "Total",
  num_of_games = sum(joined_tbl$num_of_games),
  num_of_observations = sum(joined_tbl$num_of_observations))

tbl2 <- bind_rows(joined_tbl, totals_row) # Append totals

tbl2 %>% kable(col.names = c("Day of Week", "Games", "Observations"))
```

\newpage

## Distribution of Offenses on Saturdays
```{r distribution-of-offenses-on-saturdays}
# Figure showing the mean number of offenses (assaults or vandalisms) on Saturdays 
# (with standard errors), broken out by whether a home game, away game, 
# or no game was played. 
final_table %>% 
    mutate(day_of_the_week = wday(day, label = TRUE)) %>%
    filter(day_of_the_week == "Sat") %>% 
    group_by(game_status) %>%
    summarize(avg_offenses = mean(assault + vandalism), 
              std_error = sd(assault + vandalism) / sqrt(n())) %>%
    ggplot(aes(x = game_status, y = avg_offenses)) +
    geom_errorbar(aes(ymin = avg_offenses - std_error, ymax = avg_offenses + std_error), width = 0.4) +
    geom_point(size = 5, color = "blue") +
    labs(x = "Game Status",
         y = "Average Number of Offenses", 
         title = "Mean Number of Offenses on Saturdays") + 
    scale_x_discrete(labels = c("no_game" = "No Game",
                                "home_game" = "Home Game",
                                "away_game" = "Away Game"))
```

\newpage

## Distribution of Offenses on Saturdays By Team
```{r distribution-of-offenses-on-saturdays-per-team}
# Per-team plot: Show the same comparison separately for each college town / team instead 
# of aggregated across all teams.
final_table %>% 
    mutate(day_of_the_week = wday(day, label = TRUE)) %>%
    filter(day_of_the_week == "Sat") %>% 
    group_by(game_status, team_name) %>%
    summarize(avg_offenses = mean(assault + vandalism), 
              std_error = sd(assault + vandalism) / sqrt(n())) %>%
    ggplot(aes(x = game_status, y = avg_offenses)) +
    geom_errorbar(aes(ymin = avg_offenses - std_error, ymax = avg_offenses + std_error), width = 0.4) +
    geom_point(size = 3, color = "blue") +
    facet_wrap(~team_name) +
    labs(x = "Game Status",
         y = "Average Number of Offenses", 
         title = "Mean Number of Offenses on Saturdays Per Team") + 
    scale_x_discrete(labels = c("no_game" = "No Game",
                                "home_game" = "Home Game",
                                "away_game" = "Away Game"))
```

\newpage

## Appendix: Descriptive Statistics for Count Variables
```{r replicate-appendix}
# function to get individual stats
quantile_generator <- function(table, type){
    select(table,type) %>%
    summarise(mean = mean(.data[[type]]),
        std = sd(.data[[type]]),
        Q1 = quantile(.data[[type]], 0.25),
        Q2 = quantile(.data[[type]], 0.50),
        Q3 = quantile(.data[[type]], 0.75),
        Q4 = quantile(.data[[type]], 0.90))
}

types = c("assault", "vandalism") # type of offences

# mapping out from our funciton to collect stats
all_day <- map_dfr(types,
                   quantile_generator,
                   table = final_table) # All Day stats

home_game <- map_dfr(types,
                     quantile_generator,
                     table = final_table %>% 
                     filter(game_status == "home_game")) # Home Game stats

away_game <- map_dfr(types,
                     quantile_generator,
                     table = final_table %>% 
                     filter(game_status == "away_game")) # Away Game stats

no_game <- map_dfr(types,
                   quantile_generator,
                   table = final_table %>% 
                   filter(game_status == "no_game")) # No Game stats

# binding all stats table into final appendix table
appendix <- bind_rows(all_day %>% mutate(day_type = "All Days", offense = types) ,
                      no_game %>% mutate(day_type = "No Game", offense = types),
                      home_game %>% mutate(day_type = "Home Game", offense = types) ,
                      away_game %>% mutate(day_type = "Away Game", offense = types))

# Formating table
appendix <- appendix %>% 
    pivot_longer(-c(day_type, offense),
                 names_to = "Stats", 
                 values_to = "Value") %>% 
                 pivot_wider(names_from = day_type, 
                             values_from = Value)
                            
appendix %>% kable()
```

\newpage

## Table 3: College Football Games and Number of Offenses

### Linear Model
```{r linear-model-table3}
# Linear model of daily offense counts on home-game and away-game indicators, 
# controlling for agency fixed effects, day of week, month, and year.
regression_df <- final_table %>%
    mutate(day_of_the_week = factor(wday(day, label = TRUE), ordered = FALSE), # fixed effect : day of week
           month = factor(month(day), ordered = FALSE), # fixed effect : month of the year
           year = factor(year(day), ordered = FALSE)) %>%  # fixed effect : year of the season
    mutate(home_game = case_when(game_status == "home_game" ~1, TRUE ~0), # Adding boolean column for home_game 
           away_game = case_when(game_status == "away_game" ~1,TRUE  ~0)) # Addin boolean column for away_game

# Linear regression model for predicting number of assault
assaults_model <- lm(assault ~ 
                    home_game + # effect of home_game
                    away_game + # effect of away_game
                    day_of_the_week + # fixed effect : day of week
                    month + # fixed effect : month 
                    year +  # fixed effect : year
                    factor(ori, ordered = FALSE), #fixed effect : ORI's/towns
                    data = regression_df )

# Linear regression model for predicting number of vandalism
vandalism_model <- lm(vandalism ~ 
                    home_game + # effect of home game
                    away_game + # effect of away game
                    day_of_the_week + # fixed effect: day of week
                    month + # fixed effect: month 
                    year +  # fixed effect: year
                    factor(ori,ordered = FALSE), # fixed effect: ORI's/towns
                    data = regression_df)

# Create a formatted regression table using the two negative binomial models
modelsummary(
  # List the models that will appear as columns in the table
  list("Assaults" = assaults_model,
       "Vandalism" = vandalism_model),

  # Rename variables for display
  coef_map = c("home_game" = "Home game",
               "away_game" = "Away game"),

  # Hide coefficients we're not interested in displaying
  coef_omit = "Intercept|day_of_the_week|month|year|ori",

  # Add significance stars
  stars = TRUE,

  # Customize the goodness-of-fit statistics shown at the bottom
      # raw = statistic name used internally
      # clean = label displayed in the table
      # fmt = number of decimal places
  gof_map = tribble(~raw,      ~clean,                 ~fmt,
                    "nobs",    "Observations",          0),
    
  add_rows = tibble(
    # Names of additional rows (to match the paper)
    term = c("Agency fixed effects",
             "Groups"),

    # Values to display under the Assaults column
    Assaults = c("Yes", # Agency fixed effects were included
                 as.character(n_distinct(regression_df$ori))), # Number of agencies (ORIs)

    # Values to display under the Vandalism column
    Vandalism = c("Yes", # Agency fixed effects were included
                  as.character(n_distinct(regression_df$ori))) # Number of agencies (ORIs)
  ),

  title = "College Football Games and Number of Offenses"
)

# We did not get statistically significant result for effect of home_game or away_game in predicting assaults and vandalism
```

\newpage

```{r linear-p-val-table3, include=FALSE}
# Linear Assault
summary(assaults_model)$coefficients[c("home_game", "away_game"), 4]
# Linear Vandalism
summary(vandalism_model)$coefficients[c("home_game", "away_game"), 4]
```

### Negative Binomial Model
```{r negative-binomial-table3}
# Since the linear models did not reproduce the paper's findings, we fit negative binomial models 
# to better match the methodology used in the original paper.

# Negative binomial model for prediciting vandalism
vandalism_model_nb <- glm.nb(vandalism ~ 
                             home_game + # effect of home game
                             away_game + # effect of away game
                             day_of_the_week + # fixed effect: day of week
                             month + # fixed effect: month 
                             year +  # fixed effect: year
                             factor(ori,ordered = FALSE), # fixed effect: Ori's/towns
                             data = regression_df)

# Negative binomial model for assault prediction
assaults_model_nb <- glm.nb(assault ~ 
                            home_game + # effect of home_game
                            away_game + # effect of away_game
                            day_of_the_week + # fixed effect : day of week
                            month + # fixed effect : month 
                            year +  # fixed effect : year
                            factor(ori, ordered = FALSE), #fixed effect : ORI's/towns
                            data = regression_df)

# Create a formatted regression table using the two negative binomial models
modelsummary(
  # List the models that will appear as columns in the table
  list("Assaults" = assaults_model_nb,
       "Vandalism" = vandalism_model_nb),

  # Rename variables for display
  coef_map = c("home_game" = "Home game",
               "away_game" = "Away game"),

  # Hide coefficients we're not interested in displaying
  coef_omit = "Intercept|day_of_the_week|month|year|ori",

  # Add significance stars
  stars = TRUE,

  # Customize the goodness-of-fit statistics shown at the bottom
      # raw = statistic name used internally
      # clean = label displayed in the table
      # fmt = number of decimal places
  gof_map = tribble(~raw,      ~clean,                 ~fmt,
                    "nobs",    "Observations",          0,
                    "logLik",  "Log likelihood",        2),

  add_rows = tibble(
    # Names of additional rows (to match the paper)
    term = c("Agency fixed effects",
             "Groups"),

    # Values to display under the Assaults column
    Assaults = c("Yes", # Agency fixed effects were included
                 as.character(n_distinct(regression_df$ori))), # Number of agencies (ORIs)

    # Values to display under the Vandalism column
    Vandalism = c("Yes", # Agency fixed effects were included
                  as.character(n_distinct(regression_df$ori))) # Number of agencies (ORIs)
  ),

  title = "College Football Games and Number of Offenses"
)

```

\newpage

```{r negative-binomial-p-val-table3, include=FALSE}
# NB Assault
summary(assaults_model_nb)$coefficients[c("home_game", "away_game"), 4]
# NB Vandalism
summary(vandalism_model_nb)$coefficients[c("home_game", "away_game"), 4]
```

### Conclusion
Comparing our linear regression results to the original table from the paper, neither the home-game nor away-game 
coefficients were statistically significant predictors of assaults or vandalism in the linear models. 
In contrast, the original paper reported that home games were associated with approximately a 9% increase 
in assaults and an 18% increase in vandalism, both with p-values below 0.01.

To further investigate this discrepancy, we fit a negative binomial model, following the methodology 
described in the original paper. The results showed that home games were associated with an estimated 
11.3% increase in assaults and a 12.9% increase in vandalism. However, only the vandalism effect 
was statistically significant at the alpha = 0.05 level (p = 0.034), while the assault effect was borderline 
significant (p = 0.055). Away games were not statistically significant predictors in either model. Overall, 
these findings partially support the paper's conclusion that home games are associated with increased 
offense rates, although we were unable to fully replicate the magnitude and statistical significance of the 
original results.

\newpage

## Table 4: Winning Versus Losing

### Linear Model
```{r linear-model-table4}
# Extracting cols to get factors for wins and losses
regression_df <- regression_df %>% 
    mutate(home_game_loss = case_when((home_game == 1 & home_win == 0) ~1, TRUE ~0), # Adding boolean column for home_game_loss
           away_game_loss = case_when((away_game == 1 & away_win == 0) ~1,TRUE  ~0), # Addin boolean column for away_game_loss
           home_game_win = case_when((home_game == 1 & home_win == 1) ~1, TRUE ~0), # Adding boolean column for home_game_win
           away_game_win = case_when((away_game == 1 & away_win == 1) ~1,TRUE  ~0)) # Adding boolean column for away_game_win

# Linear regression model for predicting number of assaults
assaults_model_with_win_loss <- lm(assault ~ 
                                   home_game_win + # effect of home game win at home's ori
                                   home_game_loss + #effect of home game loss at home's Ori
                                   away_game_win + # effect of away game win at away's ori
                                   away_game_loss + #effect of away game loss at away's ori
                                   day_of_the_week + # fixed effect : day of week
                                   month + # fixed effect : month 
                                   year +  # fixed effect : year
                                   factor(ori, ordered = FALSE), #fixed effect : ORI's/towns
                                   data = regression_df)

# Linear regression model for predicting number of vandalism
vandalism_model_with_win_loss <- lm(vandalism ~ 
                                    home_game_win + # effect of home game win at home's ori
                                    home_game_loss + #effect of home game loss at home's Ori
                                    away_game_win + # effect of away game win at away's ori
                                    away_game_loss + #effect of away game loss at away's ori
                                    day_of_the_week + # fixed effect: day of week
                                    month + # fixed effect: month 
                                    year +  # fixed effect: year
                                    factor(ori,ordered = FALSE), # fixed effect: ORI's/towns
                                    data = regression_df)

# Create a formatted regression table using the two negative binomial models
modelsummary(
  # List the models that will appear as columns in the table
  list("Assaults" = assaults_model_with_win_loss,
       "Vandalism" = vandalism_model_with_win_loss),

  # Rename variables for display
  coef_map = c("home_game_win" = "Home game win",
               "home_game_loss" = "Home game loss",
               "away_game_win" = "Away game win",
               "away_game_loss" = "Away game loss"),

  # Hide coefficients we're not interested in displaying
  coef_omit = "Intercept|day_of_the_week|month|year|ori",

  # Add significance stars
  stars = TRUE,

  # Customize the goodness-of-fit statistics shown at the bottom
      # raw = statistic name used internally
      # clean = label displayed in the table
      # fmt = number of decimal places
  gof_map = tribble(~raw,      ~clean,                 ~fmt,
                    "nobs",    "Observations",          0),
                    
  add_rows = tibble(
    # Names of additional rows (to match the paper)
    term = c("Agency fixed effects",
             "Groups"),

    # Values to display under the Assaults column
    Assaults = c("Yes", # Agency fixed effects were included
                 as.character(n_distinct(regression_df$ori))), # Number of agencies (ORIs)

    # Values to display under the Vandalism column
    Vandalism = c("Yes", # Agency fixed effects were included
                  as.character(n_distinct(regression_df$ori))) # Number of agencies (ORIs)
  ),

  title = "Winning Versus Losing"
)
```

\newpage

```{r linear-p-val-table4, include=FALSE}
# Linear Assault
summary(assaults_model_with_win_loss)$coefficients[c("home_game_win", "home_game_loss", "away_game_win", "away_game_loss"), 4]
# Linear Vandalism
summary(vandalism_model_with_win_loss)$coefficients[c("home_game_win", "home_game_loss", "away_game_win", "away_game_loss"), 4]
```

### Negative Binomial Model
```{r negative-binomial-table4}
# Negative binomial model for prediciting vandalism
vandalism_model_with_win_loss_nb <- glm.nb(vandalism ~ 
                                           home_game_win + # effect of home game win at home's ori
                                           home_game_loss + #effect of home game loss at home's Ori
                                           away_game_win + # effect of away game win at away's ori
                                           away_game_loss + #effect of away game loss at away's ori
                                           day_of_the_week + # fixed effect: day of week
                                           month + # fixed effect: month 
                                           year +  # fixed effect: year
                                           factor(ori,ordered = FALSE), # fixed effect: ORI's/towns
                                           data = regression_df)

# Negative binomial model for assault prediction
assaults_model_with_win_loss_nb <- glm.nb(assault ~ 
                                          home_game_win + # effect of home game win at home's ori
                                          home_game_loss + #effect of home game loss at home's Ori
                                          away_game_win + # effect of away game win at away's ori
                                          away_game_loss + #effect of away game loss at away's ori
                                          day_of_the_week + # fixed effect : day of week
                                          month + # fixed effect : month 
                                          year +  # fixed effect : year
                                          factor(ori, ordered = FALSE), #fixed effect : ORI's/towns
                                          data = regression_df)

# Create a formatted regression table using the two negative binomial models
modelsummary(
  # List the models that will appear as columns in the table
  list("Assaults" = assaults_model_with_win_loss_nb,
       "Vandalism" = vandalism_model_with_win_loss_nb),

  # Rename variables for display
  coef_map = c("home_game_win" = "Home game win",
               "home_game_loss" = "Home game loss",
               "away_game_win" = "Away game win",
               "away_game_loss" = "Away game loss"),

  # Hide coefficients we're not interested in displaying
  coef_omit = "Intercept|day_of_the_week|month|year|ori",

  # Add significance stars
  stars = TRUE,

  # Customize the goodness-of-fit statistics shown at the bottom
      # raw = statistic name used internally
      # clean = label displayed in the table
      # fmt = number of decimal places
  gof_map = tribble(~raw,      ~clean,                 ~fmt,
                    "nobs",    "Observations",          0,
                    "logLik",  "Log likelihood",        2),

  add_rows = tibble(
    # Names of additional rows (to match the paper)
    term = c("Agency fixed effects",
             "Groups"),

    # Values to display under the Assaults column
    Assaults = c("Yes", # Agency fixed effects were included
                 as.character(n_distinct(regression_df$ori))), # Number of agencies (ORIs)

    # Values to display under the Vandalism column
    Vandalism = c("Yes", # Agency fixed effects were included
                  as.character(n_distinct(regression_df$ori))) # Number of agencies (ORIs)
  ),

  title = "Winning Versus Losing"
)
```

\newpage

```{r negative-binomial-p-val-table4, include=FALSE}
# NB Assault
summary(assaults_model_with_win_loss_nb)$coefficients[c("home_game_win", "home_game_loss", "away_game_win", "away_game_loss"), 4]
# NB Vandalism
summary(vandalism_model_with_win_loss_nb)$coefficients[c("home_game_win", "home_game_loss", "away_game_win", "away_game_loss"), 4]
```

### Conclusion
Comparing our linear regression results to the original paper, none of the home-game or away-game 
outcome variables were statistically significant predictors of assaults or vandalism. In contrast, 
the original study reported significant effects for both home-game wins and home-game losses.

To further investigate this discrepancy, we fit negative binomial models that matched the methodology 
used in the original paper. Although some p-values decreased in the negative binomial models, none 
reached statistical significance at the alpha = 0.05 level. Therefore, we were unable to replicate the 
paper's finding that home-game outcomes significantly affect assault and vandalism rates. Consistent 
with the original paper, away-game outcomes did not appear to have a statistically significant effect.